Pyro -Velocity: Probabilistic RNA Velocity inference from single-cell data

### 2024-08-20 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@

\caption{\label{fig-model-results}Pyro -Velocity outperforms other
RNA-velocity methods and accurately predicts cell lineage trajectories.
a,b. RNA Velocity model comparisons from fully mature peripheral blood
mononuclear cells (PBMCs, t-SNE embedded plots, a) or E15.5 mouse
pancreas (UMAP embedded plots, b-c). From left to right in panels a and
b: 1. Cells colored by cell type; 2. Stream plot of velocity vector
field defined by scVelo; 3. Stream plot of the posterior mean vector
field from Pyro -Velocity; 4. Pyro -Velocity visualization of cell
dynamics for 3 arbitrarily selected single cells; 5. Single-cell vector
field uncertainty where each cell is colored based on the scaled angular
standard deviation to visualize uncertainty based on n=30 iterated,
posterior samples depicted as arrows noted throughout our analysis; 6.
Averaged vector field uncertainty where each vector is colored by the
averaged vector field uncertainty. The scaled angular standard deviation
range is {[}0 360{]}, where 0 corresponds to the highest confidence for
a single direction, 360 corresponds to the highest uncertainty, and no
directionality preference based on 30 posterior samples (4-6). Vectors
were projected into the UMAP space using the scVelo API based on cosine
similarity-derived transition matrix. c.~UMAP embedded cells colored by
the average cell shared time across 30 posterior samples. The Spearman's
correlation between Pyro -Velocity and Cytotrace is provided.
d.~Pyro -Velocity automatic gene selection and ranking for predicting
important genes involved in pancreas development. e. Phase portraits of
posterior samples average spliced and unspliced gene expression (left),
rainbow plots showing denoised spliced gene expression (middle), and
gene expression within UMAP embedded space. Single dots correspond to
individual cells (left, right) or vertical lines (middle). Cell type
annotation coloring is the same as in panel b.}

@@

\caption{\label{suppfig-model-2-result}Extension of
Fig.~\ref{fig-model-results}, Pyro -Velocity (Model 2) outperforms
scVelo on PBMC and pancreas datasets. Same format as in
Fig.~\ref{fig-model-results} and
Supplementary Figure~\ref{suppfig-model-results}}

@@

\caption{\label{suppfig-lineage-tracing-2}Extension of
Fig.~\ref{fig-lineage-tracing} presenting an extended comparison of
Pyro -Velocity and scVelo on scRNA-seq datasets with lineage barcoding
information (LARRY dataset). a. uni-fate Monocyte lineage analysis with
Pyro -Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
Pyro -Velocity Model 2, c.~Neutrophil and Monocyte bi-fate lineage
analysis with Pyro -Velocity Model 1, and d.~multi-fate lineage analysis
with Pyro -Velocity Model 1. The panel descriptions are the same as in
Fig.~\ref{fig-lineage-tracing}}

>>>>

In conclusion, Pyro-Velocity provides the first probabilistic,
interpretable, end-to-end Bayesian inference framework for RNA Velocity
that learns cell dynamics and recovers cell fate from multimodal raw
read counts. This approach provides uncertainty estimations of cell fate
trajectories and novel visualization approaches to mitigate false
discoveries. Pyro-Velocity recovered shared time and vector fields that
are recapitulated in true developmental order and outperformed
state-of-the-art methods like Cytotrace and scVelo
(Supplementary Figure~\ref{suppfig-lineage-tracing}a).

\section{Software availability}\label{software-availability}

We developed our Pyro-Velocity models with Pyro (version 1.6.0)
probabilistic programming language and PyTorch (version 1.8.1).
Pyro-Velocity source code is available at
\url{https://github.com/pinellolab/pyrovelocity}. The version of
Pyro-Velocity (v0.1.0) used to generate the results presented in this
manuscript has been deposited on Zenodo:
\url{https://zenodo.org/record/7072876\#.Yx-2e-zMLxg}.

\section{Data availability}\label{data-availability}

All the data presented in this manuscript were collected from previously
published studies. The fully mature PBMC, pancreas datasets are
downloaded and loaded through the scVelo using
\texttt{scvelo.datasets.pancreas()}, and
\texttt{scvelo.datasets.pbmc68k()}. The LARRY inDrop sequencing data was
processed by the LARRY authors using dropEst \citep{Petukhov2018-nr} and
velocyto. This dataset can be accessed at
\url{https://figshare.com/articles/dataset/larry_invitro_adata_sub_raw_h5ad/20780344}.

\section{Acknowledgments}\label{acknowledgments}

This project has been made possible in part by grant number 2022-249212
from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon
Valley Community Foundation. L.P. received support from the U.S. NIH
(R35 HG010717) and Silicon Valley Community Foundation (2022-249212).
D.M.L is supported by R01CA226926, R01 CA215118, R01CA154923, R24
OD031955, CureSearch Acceleration Initiative, and the MGH Scholars
Award. We thank the Pinello and Langenau labs members for helpful
discussions.

\section{Author contributions}\label{author-contributions}

L.P., Q.Q., and E.B. conceived the work. Q.Q. and E.B created the
software. Q.Q. analyzed the data, and prepared the figures and tables.
L.P and D.L. supervised the project. L.P., D.L., E.B., G.L.M helped in
designing experiments and analyses. All the authors contributed to the
writing of the manuscript.

\section{Competing financial interests
statement}\label{competing-financial-interests-statement}

L.P. has financial interests in Edilytics, Inc., Excelsior Genomics, and
SeQure Dx, Inc.~L.P.'s interests were reviewed and are managed by
Massachusetts General Hospital and Partners HealthCare in accordance
with their conflict of interest policies.

@@

\caption{\label{fig-model-results}Pyro -Velocity outperforms other
RNA-velocity methods and accurately predicts cell lineage trajectories.
a,b. RNA Velocity model comparisons from fully mature peripheral blood
mononuclear cells (PBMCs, t-SNE embedded plots, a) or E15.5 mouse
pancreas (UMAP embedded plots, b-c). From left to right in panels a and
b: 1. Cells colored by cell type; 2. Stream plot of velocity vector
field defined by scVelo; 3. Stream plot of the posterior mean vector
field from Pyro -Velocity; 4. Pyro -Velocity visualization of cell
dynamics for 3 arbitrarily selected single cells; 5. Single-cell vector
field uncertainty where each cell is colored based on the scaled angular
standard deviation to visualize uncertainty based on n=30 iterated,
posterior samples depicted as arrows noted throughout our analysis; 6.
Averaged vector field uncertainty where each vector is colored by the
averaged vector field uncertainty. The scaled angular standard deviation
range is {[}0 360{]}, where 0 corresponds to the highest confidence for
a single direction, 360 corresponds to the highest uncertainty, and no
directionality preference based on 30 posterior samples (4-6). Vectors
were projected into the UMAP space using the scVelo API based on cosine
similarity-derived transition matrix. c.~UMAP embedded cells colored by
the average cell shared time across 30 posterior samples. The Spearman's
correlation between Pyro -Velocity and Cytotrace is provided. d.
Pyro -Velocity automatic gene selection and ranking for predicting
important genes involved in pancreas development. e. Phase portraits of
posterior samples average spliced and unspliced gene expression (left),
rainbow plots showing denoised spliced gene expression (middle), and
gene expression within UMAP embedded space. Single dots correspond to
individual cells (left, right) or vertical lines (middle). Cell type
annotation coloring is the same as in panel b.}

@@

\caption{\label{suppfig-model-2-result}Extension of
Fig.~\ref{fig-model-results}, Pyro -Velocity (Model 2) outperforms
scVelo on PBMC and pancreas datasets. Same format as in
Fig.~\ref{fig-model-results} and
Supplementary Figure~\ref{suppfig-model-results}.}

@@

\caption{\label{suppfig-lineage-tracing-2}Extension of
Fig.~\ref{fig-lineage-tracing} presenting an extended comparison of
Pyro -Velocity and scVelo on scRNA-seq datasets with lineage barcoding
information (LARRY dataset). a. uni-fate Monocyte lineage analysis with
Pyro -Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
Pyro -Velocity Model 2, c.~Neutrophil and Monocyte bi-fate lineage
analysis with Pyro -Velocity Model 1, and d.~multi-fate lineage analysis
with Pyro -Velocity Model 1. The panel descriptions are the same as in
Fig.~\ref{fig-lineage-tracing}.}
